---
title: "ANOVA Examples"
author: "Michelle R. Jackson"
date: "7/25/2018"
output: html_document
---

Original Date: July 11, 2018  
R version: 3.5.0 "Joy in Playing"  
Purpose: Code explains how to perform a single factor ANOVA in R  Source:http://environmentalcomputing.net/analysis-variance-single-factor/

##Analysis
Download Turtles file from URL above or Github.
```{r}
Turtles <- read.csv(file = "Turtles.csv",header=TRUE)
```
Check that the temperature variable is a factor with the 'str' "string" function.
```{r}
str(Turtles) 
```
There are numbers for the four levels of the Temperature treatment, we need to change the variable to a factor instead of an integer.
```{r}
Turtles$Temperature <- factor(Turtles$Temperature)
```
We want to create an ANOVA comparing hatching time days across temperature with the 'aov' function.
```{r}
Turtles.aov <- aov(Days ~ Temperature, data = Turtles)
```
Obtain the output of the ANOVA with the 'summary' function.
```{r}
summary(Turtles.aov)
```
Same analysis can be created with 'lm' function.
```{r}
Turtles.lm <- lm(Days ~ Temperature,data=Turtles)
```
Obtain the output of the 'lm' function.
```{r}
summary(Turtles.lm)
```
Conduct a comparison between each mean and the other means by a Tukey test with the 'TukeyHSD' function.
```{r}
TukeyHSD(Turtles.aov)
```
Look at assumptions following the code below. This line of code put two plots in the same window.
```{r}
par(mfrow = c(1,2))
```
Look at a histogram of the residuals.
```{r}
hist(Turtles.aov$residuals)
```
Look at QQplot of the data.
```{r}
plot(Turtles.aov,which=2)
```
Look at homogeneity of variance 
```{r}
plot(Turtles.aov,which=1)
```
Create a boxplot of the results.
```{r}
boxplot(Days~Temperature, data=Turtles, ylab = "Hatching time (days)", xlab = "Temperature (°C)")
```





