---
title: "Dplyr Examples"
author: "Michelle R. Jackson"
date: "7/26/2018"
output: html_document
---
Purpose: Code explains how to transform data with the 'dplyr' function
Source: R for Data Science by Garrett Grolemund & Hadley Wickham
http://r4ds.had.co.nz/transform.html 
#####Download required packages 
```{r}
if (!require("dplyr")) install.packages("dplyr")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("nycflights13")) install.packages("nycflights13") 
```
#####Analyses
Load datasets & packages.
```{r}
library(nycflights13)
library(tidyverse) 
```
See flights tibble.
```{r}
nycflights13::flights 
```
See flights overall.
```{r}
View(flights)
```
#####Dplyr & Tibble components:

* 'int'= "integers"
* 'dbl' = "doubles" (real numbers)
* 'chr' = "character vectors" (strings)
* 'dttm' = "date-times" (a date + a time)
* 'lgl' = "logical" (vectors that are only TRUE or FALSE)
* 'fctr' = "factors" (represent categorical variables w/ fixed values)
* 'date' =  "dates"

#####Dplyr basics include the following:
* (filter()) picks observations by value 
* (arrange()) reorders rows 
* (select()) picks variables by name
* (mutate()) creates new variables w/ functions from existing variables
* (summarise()) collapses many values down to a single summary

Example of looking at a subset of observations based values:
```{r}
filter(flights, month == 1, day == 1) 
```
Note: if you want to save the result from above you need the "<-" modifier 
```{r}
jan1 <- filter(flights, month == 1, day == 1) 
```
Example of printing out results & saving them as a variable:
```{r}
(dec25 <- filter(flights, month == 12, day == 25)) 
```
Example of needing "==" for equality instead of "="
```{r}
filter(flights, month == 1)
```
Use "near()" for approximations though instead of "=="
```{r}
near(sqrt(2) ^ 2,  2)
```
Additional example:
```{r}
near(1 / 49 * 49, 1)
```
Example of finding all flights that occurred in November & December:
```{r}
filter(flights, month == 11 | month == 12)
```
This code uses filter() to select every row where x is one of the values in y
```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```
Example of finding fights that weren’t delayed (on arrival or departure) 
by more than two hours. Both filters work!
```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```
To determine if a value is missing use is.na():
An example: let x be Mary's age. We don't know how old she is.
```{r}
x <- NA
is.na(x)
```
Note: filter() only includes rows when condition = TRUE; it exclude FALSE & NA
values. Want to preserve missing values? Ask for them like this:
```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```
arrange() works like filter() but it doesn't select rows, it reorders them.
```{r}
arrange(flights, year, month, day)
```
desc() reorders by a column in descending order - see:
```{r}
arrange(flights, desc(dep_delay))
```
Example of how missing values are always sorted at the end:
```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```
select() let's you closely examine a subset of your data with operations 
based on variable names.
```{r}
select(flights, year, month, day)
select(flights, year:day)
select(flights, -(year:day))
```
Examples of other helper functions you can use within the select() function:

*  starts_with("abc"): matches names beginning with “abc”
*  ends_with("xyz"): matches names ending with “xyz”
*  contains("ijk"): matches names containing “ijk”
*  matches("(.)\\1"): selects variables matching a regular expression;
  specifically variables containing repeated characters. 
*  num_range("x", 1:3): matches x1, x2 and x3.
* use rename(): renames variables
Example of rename() function
```{r}
rename(flights, tail_num = tailnum)
```
Using select() with the everything() helper to move variables 
to the start of the data frame.
```{r}
select(flights, time_hour, air_time, everything())
```
Note: mutate() adds new columns at the end of your dataset.
To see all the columns use View().
Example of creating a narrower data set and changes with mutate.
```{r}
flights_sml <- select(flights, 
                      year:day, 
                      ends_with("delay"), 
                      distance, 
                      air_time
)
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       speed = distance / air_time * 60
)
mutate(flights_sml,
       gain = dep_delay - arr_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours
)
```
To keep the new variables from the code above use transmute():
```{r}
transmute(flights,
          gain = dep_delay - arr_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours
)
```
summarise() collapses a data frame to a single row:
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```
summarise() works best with group_by() to go from analyzing the complete dataset
to individual groups (e.g., a data frame grouped by date gives
you average delay per date)
```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```
Example of exploring the relationship between the distance and average flight
delay for each location:
```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```
Guided example from this chapter about how to prepare this data for additional
analyses:

* Group flights by destination
* Summarise to compute distance, average delay, and number of flights
* Filter to remove noisy points and Honolulu airport (which is an outlier
)
To deal with the tediousness of having to name each intermediate data frame use
the pipe, %>%:
```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```
 %>% when reading code is pronounced “then”
Examples: x %>% f(y) makes f(x, y)
x %>% f(y) %>% g(z) makes g(f(x, y), z)

Additional example of using %>% but the issue of missing values comes up:
```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```
To remove missing values use the na.rm argument:
```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))
```
How to remove cancelled flights:
```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```
The authors reccommend when doing an aggregation like above, 
it’s helpful include either a count (n()), 
or a count of non-missing values (sum(!is.na(x)))
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )
ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)
```
How to create a scatterplot of number of flights vs. average delay:
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```
Here's how to filter out the groups with the smallest numbers of observations, 
and integrate ggplot2 into dplyr flows:
```{r}
delays %>% 
  filter(n > 25) %>% 
  ggplot(mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

####Useful 'hot keys' reccommended by the authors

*  Cmd/Ctrl + Enter: sends the whole block of code  
*  Cmd/Ctrl + Shift + P : resends previously sent chunk from editor to console
  enabling editing of that intial block

See other great examples of how to work with dplyr here at the link above!