---
title: "TidyR Examples"
author: "Michelle R. Jackson"
date: "7/26/2018"
output: html_document
---
Original Date: July 11, 2018 
R version: 3.5.1 "Feather Spray"  
Purpose: Code explains how to clean and organize data with the 'tidyr'/'tivdyverse' packages.  
Note: According to Hadley Wickham (author of 'tidyr' and 'tidyverse' packages)  
datasets are outlined as such:

* Each variable must have its own column.
* Each observation must have its own row.
* Each value must have its own cell.

#####Source:http://garrettgman.github.io/tidying/

#####Source:http://r4ds.had.co.nz/tidy-data.html

#####Download required packages
```{r}
if (!require("tidyr")) install.packages("tidyr")
if (!require("devtools")) install.packages("devtools")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
```


#####Analyses
Load 'tidyr'package to access functions to help tidy data
```{r}
library(tidyr)
```

load 'devtools' package to download data and packages from online developers
```{r}
library(devtools)
```
Download DSR dataset from Garrett Grolemund's Github
```{r}
devtools::install_github("garrettgman/DSR") 
```
Load DSR dataset that you just downloaded
```{r}
library(DSR)
```
Load 'tidyverse' package for access to additional packages to help tidy data
Also load 'dplyr' package - for more information on dplyr see tutorial on Github
```{r}
library(tidyverse)
library(dplyr)
```
Show tables 1-5 to see if they are tidy or need to be "cleaned"
```{r}
table1 
table2
table3 
table4
table4a
table4b
table5
```
Only table 1 is currently tidy!

###Example of working with tidy data by computing rate per 10,000
```{r}
table1 %>% 
  mutate(rate = cases / population * 10000)
```
Compute cases per year
```{r}
table1 %>% 
  count(year, wt = cases)
```
Visualise changes over time
```{r}
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```
Issues one might encounter with untidy data include: 
* one variable might be spread across multiple columns
* one observation might be scattered across   multiple rows 

Let's fix the column names in table 4a (e.g., 1999 & 2000) using 'gather' 
```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```
* key = the name of the variable whose values for the column names
* value = the name of the variable whose values spread over the cells 

code to tidy table 4b
```{r}
table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population") 
```
Combine table 4a & 4b into 1 tibble
```{r}
left_join(table4a, table4b)
```
* spread - a function used when an observation is scattered 
  across multiple rows (e.g., table 2)
```{r}
table2
table2 %>%
  spread(key = "key", value = "value")
```
Other issues include one column containing two varables (e.g., table 3) - 
this can be fixed with the 'seperate' function
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```
you can use a specific character to seperate your column (e.g., "/" below )
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```
you can improve the default behavior of 'seperate" using "convert=TRUE" 
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```
you can pass the vector of integers to "sep" 
```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```
Use the 'unite' function - it combines multiple columns into a single column 
Try it on tidyr::table5 - 'sep' remove the underscore in the year column
```{r}
tidyr::table5 %>% 
  unite(new, century, year)
tidyr::table5 %>% 
  unite(new, century, year, sep = "")
```
how to fix missing data
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks %>% 
  spread(year, return) %>% 
  gather(year, return, `2015`:`2016`, na.rm = TRUE)
stocks %>% 
  complete(year, qtr)
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment %>% 
  fill(person)
```
Case study with all tidyr examples using WHO TB data
```{r}
who
who1 <- who %>% 
  gather(new_sp_m014:new_rel_f65, key = "key", value = "cases", na.rm = TRUE)
who1
```
Look at the new structure values by using the 'count function'
```{r}
who1 %>% 
  count(key)
```
Fix inconsistency with the names with code below:
```{r}
who2 <- who1 %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))
who2
```
Now separate the values in each code like this:
```{r}
who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```
Drop inconsitent columns by doing the following:
```{r}
who3 %>% 
  count(new)
who4 <- who3 %>% 
  select(-new, -iso2, -iso3)
```
Split 'sex' and 'age' into seperate groups:
```{r}
who5 <- who4 %>% 
  separate(sexage, c("sex", "age"), sep = 1)
who5
```
Now here's all that together in one pipe!
```{r}
who %>%
  gather(key, value, new_sp_m014:new_rel_f65, na.rm = TRUE) %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

